---
title: "Chapter 5 homework"
format:
  html:
    embed-resources: true
  pdf:
    pdf-engine: xelatex
execute:
  echo: true
  message: false
  warning: false
editor: 
  markdown: 
    wrap: 72
---

## Setup

```{r}
library(tidyverse)
library(here)
library(broom)
library(pwr)
library(psych)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## General notes on the model-comparison workflow

For most questions below, you will compare:

-   **Model C (constrained / null model)**: a model where the intercept
    is fixed (often to 0 or to a known/reference value).
-   **Model A (augmented model)**: a model where the intercept is
    estimated from the data.

For a one-parameter intercept-only Model A,

$$
\text{Model A: } Y_i = b_0 + \varepsilon_i
$$

and for a fixed-intercept Model C (often $b_0$ fixed to some value $c$),

$$
\text{Model C: } Y_i = c + \varepsilon_i.
$$

The key quantities:

-   $\mathrm{SSE}$ is sum of squared errors.
-   $\mathrm{PRE} = \dfrac{\mathrm{SSE}_C - \mathrm{SSE}_A}{\mathrm{SSE}_C}$.
-   For a one-parameter comparison (typically $P_A - P_C = 1$), the
    usual model-comparison $F$ statistic is:

$$
F = \frac{(\mathrm{SSE}_C - \mathrm{SSE}_A)/(P_A - P_C)}{\mathrm{SSE}_A/(n - P_A)}.
$$

------------------------------------------------------------------------

# Part A: Hand-calculation practice (small, self-contained datasets)

## A1. Schizophrenia age of onset

Assume the mean age of onset of schizophrenia in a sample of $n=50$
persons was $\bar{y} = 21.5$ years and the variance was $s^2 = 2.5$
(years$^2$).

1.  Specify the model that uses the mean to describe the data (Model A),
    in both words and in equation form. $$
    Y_i = 21.5 + \varepsilon_i
    $$
2.  What is the $\mathrm{SSE}$ for this model?

*(Use the definitional relationships between variance, sums of squares,
and SSE for an intercept-only model.)*

```{r}
sse <- ((50 - 1) * 2.5)
sse
```

------------------------------------------------------------------------

## A2. Multiculturalism vs. colorblindness (paired data)

Ryan, Hunt, Weible, Peterson, and Casas (2007) investigated Americans’
interethnic ideologies. They asked college students to indicate the
extent to which they believed that multiculturalism and colorblind
ideologies would improve interethnic relations in the U.S.

Assume you have data for 10 Black participants:

| Subjid |  MC |  CB | Diff (MC − CB) |
|-------:|----:|----:|---------------:|
|      1 | 5.0 | 4.0 |            1.0 |
|      2 | 4.0 | 3.5 |            0.5 |
|      3 | 6.0 | 4.0 |            2.0 |
|      4 | 6.0 | 6.0 |            0.0 |
|      5 | 5.0 | 4.5 |            0.5 |
|      6 | 4.5 | 5.0 |           -0.5 |
|      7 | 5.0 | 3.0 |            2.0 |
|      8 | 6.5 | 4.5 |            2.0 |
|      9 | 7.0 | 5.0 |            2.0 |
|     10 | 6.5 | 3.0 |            3.5 |

### A2a. By hand

1.  Compute **by hand** the mean, variance, and standard deviation of the difference variable. Use definitional formulas.

Mean = 1.3 Variance = 1.456 Standard deviation = 1.21

2.  Compute $\mathrm{SSE}_C$ for the simple model that predicts no
    difference (i.e., that the difference is 0). This is Model C.

SSE_C = 30

3.  Compute $\mathrm{SSE}_A$ for the simple model that best describes
    the differences based on the data (Model A).

SSE_A = 13.1

4.  Compute PRE and $F$.

PRE = 0.563 F = 11.611

5.  Specify the null hypothesis tested by comparing these two models.

$$\Delta_i = 0 + \varepsilon_i$$

6.  Indicate whether you would reject or fail to reject Model C / the
    null hypothesis.

PRE\* = 0.362 F\* = 5.117

7.  Write up your findings in journal style.

In this sample from Ryan, Hunt, Weible, Peterson, and Casas' (2007) study, the average Black college student believed that multiculturalism would better improve interethnic relationships in the US than colorblind ideologies, with an average difference of 1.3. Given the observed F statistic of 11.611 and the critical F statistic of 5.117, we reject the null hypothesis that there is no difference in attitudes on multiculturalism and colorblind ideologies. 

*(You can optionally use R at the end to check arithmetic, but do the
main work by hand.)*

### A2b. In R (reproducing your values)

Recreate the analysis in R and verify the quantities above (mean,
variance/SD, $\mathrm{SSE}_C$, $\mathrm{SSE}_A$, PRE, $F$, and $p$).

```{r}
subjid <- c(1:10)
mc <- c(5, 4, 6, 6, 5, 4.5, 5, 6.5, 7, 6.5)
cb <- c(4, 3.5, 4, 6, 4.5, 5, 3, 4.5, 5, 3)

df <- data.frame(subjid, mc, cb)

df <- df |>
  mutate( diff = mc - cb)

m <- df |>
  summarize(mean = mean(diff))
m

variance <- df |>
  summarize(variance = var(diff))

sd <- sqrt(variance)

df <- df |>
  mutate(diff_dev = mc - cb)

model_c <- lm(diff_dev ~ 0, data = df)
model_a <- lm(diff_dev ~ 1, data = df)

sse_c <- deviance(model_c)
sse_a <- deviance(model_a)

pre <- ((sse_c - sse_a) / sse_c)

pa <- 1
pc <- 0

n <- 10

fstat <- (pre / (pa - pc)) / ((1 - pre) / ( n - 1))

p_val <- 1 - pf(fstat, 1, 9)

m
variance
sd
pre
fstat
p_val
```

### A2c. Confidence interval

Construct a 95% confidence interval for the multiculturalism −
colorblindness difference score in this sample.

```{r}
summary(model_a)

1.3 + (2 * .3815)
1.3 - (2 * .3815)
```

0.537 \<-\> 2.063

### A2d. Power planning with `pwr`

A researcher wants to replicate the finding with a new sample of 10
participants.

1.  Use results from this sample and the `pwr` package to estimate the
    number of participants needed to replicate your observed effect (at
    $\alpha = .05$). n = 47
2.  Use the same package to estimate the number of participants needed
    to detect a **medium** effect size (according to Cohen's very
    arbitrary definition...). n = 24

```{r}
groups <- 2
obs <- 10
within_var <- var(df$diff) # estimate
between_var <- var(c(0, 1)) * ((groups - 1) / groups)  # population

power.anova.test(groups = groups,
                 between.var = between_var,
                 within.var = within_var,
                 sig.level = .05,
                 power = .8)

power.anova.test(groups = groups,
                 between.var = between_var,
                 within.var = within_var,
                 sig.level = .05,
                 power = .5)
```

------------------------------------------------------------------------

## A3. Gym BMI (one-sample mean test)

A researcher wonders whether people who go to the gym in January have
above-normal BMIs, that is, above 24.9 (upper level of “normal,”
according to the CDC). She obtains the BMI scores below for a randomly
selected group of 10 gym users:

| Participant |  BMI |
|------------:|-----:|
|           1 | 24.0 |
|           2 | 27.0 |
|           3 | 25.0 |
|           4 | 27.0 |
|           5 | 27.0 |
|           6 | 23.0 |
|           7 | 22.0 |
|           8 | 30.0 |
|           9 | 26.0 |
|          10 | 29.0 |

1.  Specify Models C and A and the null hypothesis for this statistical
    test. What is this test commonly called? 
    $$\text{Model A: } Y_i = 26 + \varepsilon_i
    $$ 
    $$
    \text{Model C: } Y_i = 24.9 + \varepsilon_i.
    $$ $$
    \text{Null Hypothesis } (H_0):  b_0 = 24.9
    $$
2.  Conduct the analysis **by hand**.
mean = 26
predicted mean = 24.9
sse for model c = 70.1
sse for model a = 58
pre = 0.173
3.  Conduct the analysis in R.
4.  Provide a statistical conclusion.
5.  Write up your findings in journal style.

```{r}
id <- c(1:10)
bmi <- c(24.0, 27.0, 25.0, 27.0, 27.0, 23.0, 22.0, 30.0, 26.0, 29.0)

d <- data.frame(id, bmi)

d <- d |>
  mutate(bmi_dev = bmi - 24.9)

m_c <- lm(bmi_dev ~ 0, data = d)
m_a <- lm(bmi_dev ~ 1, data = d)

summary(m_c)
summary(m_a)

ssec <- deviance(m_c)
ssea <- deviance(m_a)

pre_2 <- ((ssec - ssea) / ssec)

pa <- 1
pc <- 0

n_2 <- 10

fstat_2 <- (pre_2 / (pa - pc)) / ((1 - pre_2) / ( n_2 - 1))

pval <- 1 - pf(fstat_2, 1, 9)


pre_2
fstat_2
pval

qpre <- function(p,
                 df1,
                 df2) {
  
  f <- qf(p = p,
     df1 = df1,
     df2 = df2)
  
  crit_pre <- f / (f + (df2/df1))
                   
  return(crit_pre)
  
}

pre_crit <- qpre(p = .95, 1, 9)
f_crit <- qf(1 - .05, 1, 9)

pre_crit
f_crit
```
Given the critical F and pre values of 5.117 and 0.382 respectively, we fail to reject the null hypothesis that people who go to the gym in January have normal BMIs.
------------------------------------------------------------------------

# Part B: Fitness dataset (model comparison, confidence intervals, power)

## B0. Read the data

Read in the `fitness.csv` dataset from the course project `data/`
folder.

```{r}
fitness <- read_csv(here("data", "fitness.csv"))
```

------------------------------------------------------------------------

## B1. Resting pulse vs. “normal” (72 bpm)

The variable `RSTPULSE` contains resting pulse rate for each participant
in the fitness study. The normal pulse rate for men is 72.

### B1a. 95% CI for $b_0$ in Model A

Construct a 95% confidence interval for $b_0$ in Model A.

-   Do this **by hand** using a critical value (e.g.,
    $t_{\alpha/2, \, df}$), and then confirm with R.
-   Make note of what in R output corresponds to your hand calculation.

For reference, the by-hand CI for an intercept estimate is:

$$
\hat{b}_0 \pm t_{\alpha/2, \, df} \cdot SE(\hat{b}_0).
$$

```{r}
sd <- 8.294
se <- ((sd)/sqrt(31))

upper <- 53.7 + (2 * se)
lower <- 53.7 - (2 * se)

upper
lower
```

50.721 \<-\> 56.679

### B1b. Journal-style summary

Write a complete summary:

1.  briefly explain the data;
2.  set up the analysis (describe the comparison without using “Model A/C” language if you prefer);
3.  report mean (and SD), test statistic ($t$ or $F$), PRE, $p$, and the
    confidence interval; and
4.  state statistical and substantive conclusions.

```{r}
fitness <- fitness |>
  mutate(rstpulse_dev = RSTPULSE - 72)

modelc <- lm(rstpulse_dev ~ 0, data = fitness)
modela <- lm(rstpulse_dev ~ 1, data = fitness)

summary(modelc)
summary(modela)

fitness |>
  summarize(mean = mean(RSTPULSE))

c_sse <- deviance(modelc)
a_sse <- deviance(modela)

c_sse
a_sse

fit_pre <- (c_sse - a_sse) / c_sse
fit_pre

fit_true <- fit_pre / ((1) / ((1 - fit_pre) / (30)))
```

### B1c. Estimate of true PRE

If you wanted to replicate this study, what would be your best guess for
the *true* value of PRE?

```{r}
true_pre <- 1 - ((1 - fit_pre) * ((1/30)))
true_pre
```

### B1d. Power for a replication at the same $n$

Given your estimate of true PRE, if you replicated with the same sample
size, what would be your power to detect the effect? Use
`pwr.f2.test()`.

A common translation between PRE and Cohen's $f^2$ for this
model-comparison setup is:

$$
f^2 = \frac{\mathrm{PRE}}{1-\mathrm{PRE}}.
$$

```{r}
f_sqr <- (fit_pre / (1- fit_pre))
f_sqr

pwr.f2.test(1, 30, 5, sig.level = .05)
```

Power = 1

------------------------------------------------------------------------

## B2. Did running increase pulse rate? (within-person change)

Using the same dataset, compare resting pulse (`RSTPULSE`) to post-run
pulse (`RUNPULSE`). Did running increase pulse rate?

Hint: create a new variable such as:

$$
\text{change} = \text{RUNPULSE} - \text{RSTPULSE}.
$$

### B2a. 95% CI for $b_0$ in Model A

Construct a 95% CI for $b_0$ in your Model A for the change score. Do it by hand, then confirm with R.
mean = 116
variance = 120
sd = 10.9
se = 1.96

112.08 <-> 119.94
```{r}
fitness <- fitness |>
  mutate(pulse_change = RUNPULSE - RSTPULSE)

fitness |>
  summarize(mean = mean(pulse_change),
            variance = var(pulse_change),
            sd = sqrt(variance))
```

### B2b. Estimate of true PRE

If you wanted to replicate this study, what would be your best guess for the true value of PRE?

```{r}
model_0 <- lm(pulse_change ~ 0, data = fitness)
model_1 <- lm(pulse_change ~ 1, data = fitness)

sse0 <- deviance(model_0)
sse1 <- deviance(model_1)

pulse_pre <- (sse0 - sse1) / sse0
true_pulse_pre <- 1 - ((1 - pulse_pre) * ((1/30)))
true_pulse_pre
```

### B2c. Required effect size for power = .80 at n = 100

How big would true PRE need to be to yield power = .80 in a future study
with $n=100$ participants? Use `pwr.f2.test()`.

```{r}
pwr.f2.test(1, 99, sig.level = 0.05, power = 0.8)

f_2 <- 0.079

pre_pulse <- .079/1.079
pre_pulse
```

## B3. Oral contraceptives and systolic blood pressure (Rosner example)

Ten women are studied to determine if oral contraceptives increase
systolic blood pressure (mm Hg). Baseline blood pressure is measured
before and then again several months after oral contraceptives are
started. The following **changes** in systolic blood pressure are
observed:

13, 3, -1, 9, 7, 7, 6, 4, -2, 2

A summary table (from SAS) reports:

|      Mean |   Std Dev |         USS |         CSS |         T | Prob \> |
|----------:|----------:|------------:|------------:|----------:|--------:|
| 4.8000000 | 4.5655716 | 418.0000000 | 187.6000000 | 3.3246511 |  0.0089 |

Here, **USS** (“uncorrected sum of squares”) represents squared error around the model $Y_i = 0 + \varepsilon_i$ (i.e., Model C with intercept fixed at 0). **CSS** (“corrected sums of squares”) represents squared error around $Y_i = b_0 + \varepsilon_i$ (Model A with estimated intercept).

### B3a. Interpret the output using model comparison

Use this output to determine whether blood pressure significantly
increased after oral contraceptives were begun.

1.  Specify Models A & C (equations are encouraged).
2.  Specify the null hypothesis.
3.  Calculate PRE and $F$.
4.  State your statistical and substantive one-sentence conclusions.

Model C:
$$\text{RSTPULSE}_i = 0 + \varepsilon_i$$
Model A:
$$\text{RSTPULSE}_i = 4.8 + \varepsilon_i$$
Null hypothesis:
$$H_0: b_0 = 0$$

```{r}
cont_pre <- (418 - 187.6) / 418
cont_pre

f_cont <- (cont_pre / (1)) / ((1 - cont_pre) / (9))
f_cont

critical_f <- qf(.95, 1, 9)
critical_f
```
Because the F statistic is greater than the critical F value eof 5.11, we can reject the null hypothesis. 

### B3b. Estimate of true PRE

If you wanted to replicate this study, what would be your best guess for the true value of PRE?

```{r}
true_cont_pre <- 1 - ((1 - cont_pre) * ((1/9)))
true_cont_pre
```

### B3c. Sample size for power targets

What sample size would you need to achieve power = .80? What sample size would you need to achieve power = .90? Use `pwr.f2.test()`.

```{r}
pwr.f2.test(1, f2 = .901, sig.level = .05, power = .8)

pwr.f2.test(1, f2 = .901, sig.level = .05, power = .9)
```
n = 10
n = 13

------------------------------------------------------------------------

## B4. Power and design choices (GPA example)

In a particular year, a graduate program admitted 24 new students. The
admissions committee wants to interpret the students’ first-year GPA.

### B4a. Approximate power for a “moderate” effect

Graduate students in this program are required to maintain a 3.0
average. The committee plans to compare:

-   Model C: GPA fixed at 3.0
-   Model A: GPA estimated as the students’ mean

If the true difference corresponds to what the text refers to as a
“moderate” effect in the social sciences, what are the chances that the
committee will correctly conclude Model A is better than Model C if they
use $\alpha = .05$?

1.  Calculate approximate power using the book approach.
2.  Verify using `pwr` (it will not match exactly, but should be close).

```{r}
pwr.anova.test(k = 2, n = 24, f = .5, sig.level = .05)
```

### B4b. Two ways to increase power without increasing n

Given that they cannot admit additional students, state two things they could do to increase power.

An alternative way to increase power would be to increase observations of students' GPAs or to decrease the level of statistical significancy at which they seek to estimate. 

### B4c. Try-out question: setting up a conditional model comparison

The committee wants to know whether GRE-Verbal, GRE-Math, and
undergraduate GPA are useful predictors of first-year GPA.

We have not yet learned how to *test* these conditional models, but set
up the comparison:

-   Model C: predicts the mean
-   Model A: predicts GPA conditionally using all 3 predictors

Specify both models and indicate $n - P_A$ and $P_A - P_C$ for this
comparison.
$$\text{Model A: } GPA_i = b_0 + a(GREVerbal_i) + b(GREMath_i) + c(UndergradGPA) + \varepsilon_i$$ 
    $$
    \text{Model C: } GPA_i = 3.0 + \varepsilon_i.
    $$ $$
    \text{Null Hypothesis } (H_0):  b_0 = 3.0
    $$
    $n - P_A$ = 20
    $P_A - P_C$ = 3

------------------------------------------------------------------------

# Part C: Your own data (leave code blocks empty)

These questions refer to the model comparison you tested previously using your own dataset.

## C1. Confidence interval for $b_0$

Identify the 95% confidence interval for $b_0$ in Model A. Explain what this interval represents.

```{r}
gss2024 <- readRDS(file = here::here("data", "gss2024.rds"))

gss <- gss2024 |>
  select(courtsnv) |>
  drop_na() |>
  haven::zap_labels()

gss_recoded <- gss |>
  mutate(
    courtsdv = recode(courtsnv,
                        `1` = -1,
                        `2` = 1,
                        `3` = 0)
    )

gss_recoded |>
  summarize(
    n = n(),
    mean_courts = mean(courtsdv, na.rm = TRUE),
    sd_courts = sd(courtsdv, na.rm = TRUE)
  )

gss_recoded <- gss_recoded |>
  mutate(dev = courtsdv - 0)

mod_c <- lm(dev ~ 0, 
            data = gss_recoded)

mod_a <- lm(dev ~ 1, 
            data = gss_recoded)

summary(mod_c)
summary(mod_a)

sse_c_own <- deviance(mod_c)
sse_a_own <- deviance(mod_a)
pre_own <- (sse_c_own - sse_a_own) / sse_c_own
pre_own
f_own <- (pre_own / (1)) / ((1 - pre_own) / (615))
f_own
```

## C2. Journal-style summary

Write a complete journal-style summary for your data.

```{r}
critical_pre_own <- qpre(p = .95, 1, 615)
f_critical_own <- qf(.95, 1, 615)

critical_pre_own
f_critical_own
```
Given the critical F statistic and PRE value of 2.85 and 0.006 respectively, we can reject the null hypothesis that opinions are evenly distributed between "too harshly" and "not harshly enough". 
## C3. Estimate of true PRE

If you wanted to replicate this study, what would be your best guess for
the true value of PRE?

```{r}
true_pre_own <- 1 - ((1 - pre_own) * ((1/615)))
true_pre_own
```

## C4. Sample size for power = .80

Given your estimate of true PRE, what sample size would you need to
achieve power = .80? Use `pwr.f2.test()`.

```{r}
f2_own <- true_pre_own / (1 - true_pre_own)
# pwr.f2.test(u = 1, f2 = f2_own, power = .0) ## Unsure why this is resulting in an error
```

## C5. Sketch the PRE distributions

Draw a chart that shows the distribution of PRE under both the $H_0$ and
$H_1$ hypotheses. Label $\alpha$, $\beta$, power, and the surprise
point.

```{r}
# Unsure what this is asking -- will ask in class
```
