---
title: "Chapter 6 homework"
format:
  html:
    embed-resources: true
---

## Setup

```{r}
library(tidyverse)
```

## Data

This homework uses a built-in dataset from base R (`state.x77`). Run the code below to create the working tibble `states`.

```{r}
state.x77_with_names <- as_tibble(state.x77,
                                  rownames = "state")

states <- bind_cols(state.x77_with_names,
                    region = state.division) |>
  janitor::clean_names() |>
  mutate(
    south = as.integer(
      str_detect(as.character(region), "South")
    )
  )
```

Throughout, use:

- outcome: `life_exp`
- predictor: `murder`

# Question 1: Descriptive statistics

## 1a. Descriptive statistics (psych)

Use `psych::describe()` (or something else if you want) to get descriptive statistics for `life_exp` and `murder`.

## 1b. Descriptive statistics (tidyverse)

Compute (at minimum) the mean and standard deviation of `life_exp` and `murder` using `summarize()`.

# Question 2a: Regression and model comparison

You will evaluate whether `murder` predicts `life_exp`.

## 2a(i). Write the model comparison and null hypothesis

Write Model C and Model A in words and equations, and state the null hypothesis.

A typical setup for testing whether a slope is needed is:

- Model C (constrained): slope fixed at 0
- Model A (augmented): slope estimated from the data

Use the notation below (fill in the missing pieces in your write-up):

Model C:
$$
Y_i = \;\;\;\;\;\;\;\; + \varepsilon_i
$$

Model A:
$$
Y_i = \;\;\;\;\;\;\;\; + \;\;\;\;\;\;\;\; X_i + \varepsilon_i
$$

Null hypothesis (in symbols and words):
$$
H_0: \;\;\;\;\;\;\;\;
$$

## 2a(ii). Fit both models

## 2a(iii). Compute SSE(C), SSE(A), PRE, and F

Compute:

- $\mathrm{SSE}_C$ and $\mathrm{SSE}_A$ from the fitted models
- $\mathrm{PRE} = (\mathrm{SSE}_C - \mathrm{SSE}_A)/\mathrm{SSE}_C$
- the model-comparison $F$ statistic

Use `deviance()` for SSE (RSS) and `df.residual()` as needed.

$$
F = \frac{(\mathrm{SSE}_C - \mathrm{SSE}_A)/(P_A - P_C)}{\mathrm{SSE}_A/(n - P_A)}.
$$

## 2a(iv). Write the prediction equation and interpret coefficients

Write the prediction equation implied by Model A:

$$
\widehat{\text{life_exp}} = \;\;\;\;\;\;\;\; + \;\;\;\;\;\;\;\; (\text{murder})
$$

Interpret:

- the intercept
- the slope

## 2a(v). Conclusion

State:

- statistical conclusion (report $F$ or $t$, PRE, and $p$)
- substantive conclusion (in plain language)

# Question 2b: Center the predictor

Create a centered version of murder:

## 2b(i). Create `murder_c` and refit the model

## 2b(ii). Report SSE(A), SSE(C), PRE, and F again

Compute the same fit statistics (SSE, PRE, F). Compare them to the results from Question 2a.


## 2b(iii). Interpret coefficients after centering

Write the prediction equation for the centered model and interpret the intercept and slope.

In particular: what does the intercept represent *now*?


## 2b(iv). Plot and compare

Make two scatterplots with regression lines:

1. `life_exp` vs. `murder`
2. `life_exp` vs. `murder_c`

How are these plots different?

# Question 2c: Change the units of the predictor

Create a transformed predictor by standardizing `murder`.

## 2c(i). Create `murder_std` and refit the model

## 2c(ii). Report SSE(A), SSE(C), PRE, and F again

Compute SSE, PRE, and F again and compare them to 2a and 2b.

## 2c(iii). Interpret coefficients after rescaling

Write the prediction equation and interpret the slope carefully in the new units.

# Question 3: Predicted values and what transformations do (and do not) change

You now have three versions of the same basic relationship:

1. using `murder`
2. using `murder_c`
3. using `murder_std`

## 3a. Predicted life expectancy at a specific murder rate

Using each of your three fitted Model A versions, compute the predicted life expectancy for a state with `murder` = 8.

Show your work. (Be explicit about how you convert 8 into the appropriate value for `murder_c` and `murder_std`.)

## 3b. Summary paragraph

Write a brief summary explaining what you learned about transforming a predictor:

- How do these transformations affect the estimated intercept and slope?
- Do they change predicted values (when transformed consistently)?
- Do they change overall goodness of fit (SSE, PRE, $F$)?

# Project application

Do the same thing with an outcome and predictor of your choice from your own dataset. Use three transformations:

- original
- normalized X
- fully standardized

Interpret the intercepts and slopes for each of the three models. Compare the fit statistics for each of the three models.
